$\pi$SOD-M: A Methodology for Building Service-Oriented Applications in the Presence of Non-Functional Properties

### 2013-12-16 JournalToSubmit2013/metamodels.tex
Cleaned related works and section 3 ...

The {\em A-policy} based service composition ($\pi$-SCM) meta-model, described in Figure \ref{fig:e-scomposition-metamodel},
provides classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.}.
The meta-model identifies {\sc Business Collaborators}\footnote{We use {\sc capitals} for referring to meta-model classes.} and the {\sc Actions} they perform. 
Instances of this meta-model are UML activity diagrams. 
Figure~\ref{fig:e-scomposition-metamodel} show the modelling elements.
Classes pictured as coloured boxes deal with non-functional properties.
The classes pictured with a white background are those of SOD-M.

In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
\begin{itemizedTrivlist}
\item A {\sc Business Collaborator} element represents those entities that collaborate in the business processes by performing some of the required actions. 
They are graphically presented as a partition in the activity diagram. 
A collaborator can be either internal or external to the system. 
When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.

@@

\item In order to represent constraints associated to services compositions, we extended the SOD-M service composition model with two concepts: {\sc Rule} and {\sc A-policy} (see blue elements in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
An {\em A-policy} is formed by attributes and rules. 
Intuitively, the conditions of each rule will be checked.
In case of no compliance, the actions defined by the rule will be performed.
The {\sc Rule} element represents an event - condition - action rule where the {\sc Event} part represents the moment in which a constraint  will be evaluated.
An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts. 
\end{itemizedTrivlist}

>>>>

The {\em A-policy} based service composition ($\pi$-SCM) meta-model, shown in Figure \ref{fig:e-scomposition-metamodel},
provides classes to represent workflows\footnote{Workflows will be transformed into implemented service compositions.} that model  business processes.
The meta-model identifies {\sc Business Collaborators}\footnote{We use {\sc capitals} for referring to meta-model classes.} and the {\sc Actions} they perform. 
Instances of this meta-model are represented as UML activity diagrams. 
In Figure~\ref{fig:e-scomposition-metamodel} 
 Coloured boxes illustrate classes  modelling  non-functional properties and 
 white boxes represent classes modelling functional ones.

%In the meta-model of Figure~\ref{fig:e-scomposition-metamodel}:
\begin{itemizedTrivlist}
\item A {\sc Business Collaborator} element represents those entities that collaborate in the business processes by performing some of the required actions. 
They are graphically presented as a partition in the activity diagram. 
A collaborator can be either internal or external to the system. 
When the collaborator of the business is external to the system, the attribute {\sf IsExternal}\footnote{We use the {\sf sans serif} font for referring to classes defined using a meta-model.} of the collaborator is set to \textbf{true}.

@@

\item In order to represent constraints associated to services compositions, we extended introduced the concepts {\sc Rule} and {\sc A-policy} (see blue elements in the $\pi$-SCM meta-model in Figure \ref{fig:e-scomposition-metamodel}).
We model non-functional constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
An {\em A-policy} is defined by attributes and rules. 
Intuitively, the conditions of each rule will be checked.
In case of no compliance, the actions defined by the rule will be performed.
The {\sc Rule} element represents an event - condition - action rule where the {\sc Event} part represents the moment in which a constraint  will be evaluated.
An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts. 
\end{itemizedTrivlist}

### 2013-12-16 JournalToSubmit2013/modelingWithPISODM.tex
Cleaned related works and section 3 ...

As shown in Figure \ref{fig:piSOD-M}, according to MDA guidelines, $\pi$-SODM meta-models are organized into three levels: 
\begin{itemize}
\item CIM (\textit{Computational Independent Models}). Two models are defined at the CIM level: \textit{value model}
and \textit{BPMN model}. Referring to the "To Publish Music" application,  the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) 
and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).

 \item PIM (\textit{Platform Independent Models}). Our method defines four meta-models: \textit{$\pi$-UseCase}, \textit{$\pi$-ServiceProcess}, and \textit{$\pi$-ServiceCom\-po\-si\-tion}.
%and \textit{$\pi$-PEWS}.

@@

  \item PSM (\textit{Platform Specific Models}). At the PSM level we have lower-level models that can be automatically translated into actual computer programs. The \textit{$\pi$-PEWS} meta-model provides concepts for modelling service compositions.

@@

$\pi$SOD-M (Figure~\ref{fig:piSOD-M}) proposes the generation of a set of models at different levels of abstraction, as well as transformations between these models.
$\pi$SOD-M models represent both the functional aspects of the application as well as its non-functional constraints.
\textit{$\pi$-PEWS} models are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.
This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.

Extending from the highest level of abstraction of the MDA, $\pi$-SODM provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIMÕs); next,  starting from such models build platform independent models (PIMÕs) specifying the system details; next transform such models into platform specific models (PSMÕs) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system.

>>>>

$\pi$-SODM meta-models represent both the functional aspects of the application as well as its non-functional constraints.
As shown in Figure \ref{fig:piSOD-M}, according to MDA guidelines, $\pi$-SODM meta-models are organized into three levels: 
\begin{itemize}
\item CIM (\textit{Computational Independent Models}). This level focusses on the
environment of the system, as well as on its business and requirement specifications.
This viewpoint represents the software system at its highest level of abstraction.
At this stage of the development, the structure and system processing details are still unknown or undetermined.  $\pi$-SODM defines the \textit{value}
and \textit{BPMN} meta-moels for this purpose. 
%Referring to the "To Publish Music" application,  the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) 
%and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).

 \item PIM (\textit{Platform Independent Models}). This level focusses on the system functionality, hiding the details of any particular platform.
The specification defines those parts of the system that do not change from one platform to another. Our method defines four meta-models:

@@

  \item PSM (\textit{Platform Specific Models}). This level focusses on the functionality, in the context of a particular implementation platform.
Models at this level combine the platform-independent view with the specific aspects of the platform to implement the system. At the PSM level we have lower-level models that can be automatically translated into actual computer programs. The \textit{$\pi$-PEWS} meta-model provides concepts for modelling service compositions. \textit{$\pi$-PEWS} models instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{BaCAM05,Placido2010LTPD}.

@@

$\pi$-SODM (Figure~\ref{fig:piSOD-M}) proposes the generation of a set of models at different levels of abstraction, as well as transformations between these models.
This can be accomplished by defining: \textit{(i)} a model-to-model transformation, from a \textit{$\pi$-ServiceComposition} model to the corresponding PSM, and \textit{(ii)} a model-to-text transformation, from the this PSM to the composition language.

Extending from the highest level of abstraction of the MDA, $\pi$-SODM provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIMÕs); next,  starting from such models build platform independent models (PIMÕs) specifying the system details; next transform such models into platform specific models (PSMÕs) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system. 
The following sections describe the $\pi$-SODM meta-models.

### 2013-12-16 JournalToSubmit2013/relworks.tex
Cleaned related works and section 3 ...

Related works to our approach includes existing proposal dealing with the inclusion of non-functional properties in service-oriented development. Standards devoted for expressing non-functional properties for service-oriented applications are also analyzed in this section which needs to be addresses in the code generation phase. Finally exiting methodology for the development of service-oriented applications including non-functional properties are here depicted and compared with our approach.
%As a result of our study of related works, in this section we present a categorization of main concepts related the treatment if non-functional requirement in the field of service-oriented development.

\subsection{Non-functional properties in service-oriented applications}

In previous work we have carried out a systematic review \cite{placidoPhDThesis2012} in order to analyze how non-functional properties are taking into account in existing approaches for service-oriented development. Existing proposals in this field may be classified in three types of works which are described in the followed.

The first types of works are thus considering a particular non-functional concern (e.g. security) which is modeled and then "glued" or associated to functional models of an application. The work of Chollet et al. \cite{CholletL09} defines a proposal to associate non-functional quality properties (security properties in that case) to functional activities which are modeled in a web service composition model.
Schmeling et al. \cite{SchmelingCM11} present an approach and also a toolset for specifying and implementing non-functional concerns in web service compositions; non-functional concerns are modeled first in non-functional model and then related to a service composition represented in a BPMN diagram. Ovaska et al. \cite{OvaskaEHPA10} present an approach to support quality management at design time. Quality requirements are modeled in a first phase and represented then in an architectural model where quality requirements are associated to some components in the model. In our case we propose to model non-functional properties joint to the modeling of functional concerns what simplifies the web service development process. Non-functional properties are represented in our work as constraints and policies but more general quality model such as the proposed by \cite{Goeb2011,Klass2009} could be taken in consideration in further works.

Second type of works related to our proposal proposes specifics architecture or frameworks to manage and validate QoS attributes in service composition processes. The work of Xiao et al. \cite{XiaoCZBOLH08} present a framework to verify SLA compliance in composed services. Task in a BPEL process can be annotated with non-functional attributes (such as time, cost, resources, etc.) which are validated at design time. A similar work is presented in \cite{SchmelingCM11} by using a formal method that allows to choice best suited services instances according to a service composition pattern that include some QoS attributes. Babamir et al. present in \cite{Babamir2010} a framework for evaluate quality attributes of web services. By executing an algorithm, the framework allow to choice best web services for any task in a process and also candidate web service to replace original ones in cases it cannot perform at runtime. In \cite{Karunamurthy2012787} authors propose architecture in which web services may be described with non-functional characteristics that are validated in the service composition process.

Third type of works include proposal working on the evaluation and matching of non-functional properties in the process of search or invocation of particular web services.\cite{AgarwalLS09} define a policy language that provides a way to specific constraints on functional as well as non-functional properties of Web services with the aim of facilitating automated discovery and selection of services.\cite{JeongCL09,Kamalabad2012} defines also proposals to the validate quality attributes of web services when discover and compose.\cite{MohantyRP10} works also on the selection of web service for use considering non-functional properties. The paper present web services quality prediction models, which take non-functional properties into account. Similarly, \cite{Yeom2006} define a quality driven selection of web services providing a QoS model.

Our proposal by the contrary to the second and third type of proposals is focusing on the design of service-oriented application with non-functional properties and not on the validation of QoS attributes at runtime.

Other relevant works such as \cite{DAmbrogio06,Liu20121080} propose extensions to WSDL and BPEL languages respectively for the inclusion of QoS attributes that can be evaluated at runtime.

\subsection{Standards for programming non-func\-tion\-al properties in service-oriented applications}

Current standards in services' composition implement functional, non-functional constraints and communication aspects by combining different languages and protocols. WSDL and SOAP among others are languages used respectively for describing services' interfaces and message exchange protocols for calling methods exported by such services. For adding a transactional behaviour to a services' composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction. The selection of the adequate protocols for adding a specific non-functional constraints to a services' composition (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer. As a consequence, the development of an application based on a services' composition is a complex and a time-consuming process. This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications. Other works, like \cite{Fauvet05} introduce a model for transactional services composition based on an advanced transactional model.\cite{BhiriGP05} proposes an approach that consists of a set of algorithms and rules to assist designers to compose transactional services. In \cite{Vidyasankar:2004} the model introduced in \cite{SchuldtABS02} is extended to web services for addressing atomicity.

@@

@@

Beside the works presented above, there are few methodologies for service-oriented development that address the explicit modeling of non functional properties for service-based applications. Important methodologies for building service-based applications have been proposed in the last years \cite{Ramollari_asurvey,PapazoglouH06,FeuerlichtM05,soma,Arsanjani:2008}, but all of they focus mainly on the modeling and construction process of service-based business processes that represent the functionally of information systems.

Some methodologies dealing with particular non-functional properties in service orientation have also appeared. Tran et al. \cite{Tran2012531} present a model-driven approach for addressing problems related to compliance concerns. Authors define extend elements of a business process models with elements of a compliance model in which QoS policies are represented.\cite{GutierrezRF10} present PWSSec process, a methodology that provide developers with all the activities, tasks, tools, security artifacts and organizational structures necessary to design a secure WS-based solution.

@@

>>>>

A non-functional property (NFP) specifies criteria about the behavior of a system. These criteria are related to the conditions in which they are executed and to its performance. Non-functional properties are also referred as “constraints”, “quality attributes”, “quality goals”, “quality of service requirements” and “non-behavioral requirements” \cite{StGr05}. In the case of service-based applications, non-functional requirements concern the application itself as well as its component services. 
The majority of the work on NFPs uses a product-oriented approach, which is concerned with measuring how much a software system is in accordance with the set of NFPs that it should satisfy \cite{BBKL78,FePf96,KiDa96,Lyu96,MuIO90}.

These  approaches address the problem from four perspectives: (i) NFP expressed as quality of service properties (QoS attributes) that can be associated to service compositions; (ii) service matching for associating certain non-functional properties by composing specific services;  (iii) programming NFP as protocols that are asserted to service composition code; and, (iv) methodologies that provide tools for building service compositions with associated NFP.

\subsection{Classifying Non-functional properties}
We conducted a systematic review \cite{placidoPhDThesis2012} in order to analyze approaches that consider non-functional properties for developing service-oriented applications. As a result this analysis proposes a classification of NFP as a result of a study concerning software methodologies for the construction of service-oriented systems. The classification is organized in three layers representing: application modeling, services composition and services. The service composition layer serves as an integration layer between the services layer that exports methods and has associated constraints and characteristics; and the application layer that expresses requirements (NFP). At the application layer NFP can refer to business rules (e.g., only the user can publish data on her/his wall) and values for example, the email address is a string containing an “@” and a “.”. A value NFR expresses constraints about the way data and functions can be ac-cessed and executed. 
Business NFP at the service layer concerns properties that are associated to services and defines how to call their exported operations (business properties).  Value constraints concern more on the conditions in which services can be used. For example, accessing to a function within an authentication protocol.
Finally, at the service composition layer gives an abstract view of the kind of properties exported by services that can be combined for providing NFP for a composition.  For example, confidentiality, authentication, privacy and access control can provide security at the service composition layer.

\cite{ScCM10} proposes concepts and requirements for characterizing and analyzing exist-ing approaches addressing NFP for service coordinations. Therefore they propose:
\begin{itemize}
\item	A meta-model that introduces for characterizing NFP according to the entity to which they are associated: attribute, concern, action, and activity. 
\item	Six requirements for studying NFP definition approaches for service coordinations: NFP specifications, NFP actions specification, Web service subjects specification, non functional attributes execution order specification, composite Web service subjects specification, stateful non functional constraint specification.
\item	Seven requirements for studying NFP enforcement approaches: separation of con-cerns, transparent integration of functional and NFP, quantification, superimposi-tion, integration of NFP with distributed Web service, programming language in-dependence, Web service composition support.
\end{itemize}

NFP concepts are close to those defined by \cite{Souz12} but they are not organized into layers.  They correspond essentially to the service and service composition layers. The enforcement requirements describe the way NFP are weaved to service coordinations. Existing works respect all or a subset of these requirements. The chosen requirements have an impact in the way NFP are enforced at execution time.

%  Standards devoted for expressing non-functional properties for service-oriented applications are also analyzed in this section which needs to be addresses in the code generation phase. Finally exiting methodology for the development of service-oriented applications including non-functional properties are here depicted and compared with our approach.
%As a result of our study of related works, in this section we present a categorization of main concepts related the treatment if non-functional requirement in the field of service-oriented development.

\subsection{Specifying Non-functional properties}

Extensions to UML are proposed to express NFPs. NFPs can be integrated into the Class, Sequence, and Collaboration Diagrams \cite{CyPr04}. The performance engineering [DiSD02] annotates quantitative performance constraints to UML diagrams such as actor-use case communication, message in the sequence diagram, and state in the state machine diagram. The UML Profile for Modeling Quality of Service and Fault Tolerance Characteris-tics and Mechanisms \cite{Omg04} describes a set of QoS characteristics such as throughput, latency, integrity, dependability, among others. The characteristics can also be parameter-ized with for example the type of keys or types of encryption algorithms in case of the secu-rity characteristics. For all UML profiles based approaches it is generally possible to apply the stereotypes and its tagged values to arbitrary UML meta-classes, e.g. Class or Opera-tion. The ordering of stereotypes cannot be expressed but composition- related require-ments can theoretically be realized through behavioral diagrams like activity diagrams or state diagrams.

\cite{OrHe07} incorporate extra-functional properties into UML models by the use of UML stereotypes. For instance, there are stereotypes for encryption, decryption and logging which can be added to classes or operations that are representing services.
Design patterns help in documenting and communicating proven design solutions to recurring problems. Design patterns not only describe how to solve design problems but also document the designers’ intentions - why the solution is chosen over others and what tradeoffs are made.

\cite{GrYu01} observe that non-functional properties  (NFPs) are pervasive in the descriptions of design patterns and propose their systematic treatment in descrip-tions of patterns and when applying patterns during design.

Business rules provide means for expressing, managing and updating pieces of busi-ness domain knowledge independent of the rest of the application. According to the Busi-ness Rules Group \cite{Hay00}, a business rule is a statement that defines or constrains some aspect of the business. It is intended to assert business structure or to control the behavior of the business \cite{Hay00}. Business rules are usually expressed either as constraints or in the form if conditions then action. The conditions are also called rule premises. The business rule approach encompasses a collection of terms (definitions), facts (connection between terms) and rules (computation, constraints and conditional logic) \cite{Hall01}. Terms and Facts are statements that contain sensible business relevant observations, whereas rules are statements used to discover new information or guide decision-making. A business rule system is a system in which the rules are separated logically and perhaps physically from the other parts \cite{ChMe04}. \cite{ChMe04} investigate a hybrid approach, which combines business processes as known e.g., from BPEL with business rules \cite{Hay00}.

\cite{ChLa09} proposed a generative framework to address NFPs for web service composi-tions. As part of the framework they separate the orchestration meta-model from other non-functional meta-models. They propose a non-functional meta-model for security where concepts exist for requirements and how these requirements are ensured. The NFPs are then added to the composition model by adding annotations.
KAOS \cite{DaLF93,Lams01} is a goal-oriented framework for addressing functional requirements with a formal temporal logic that requirements engineer can use to precisely specify NFPs, such as performance, as part of attribute specification of functional requirements.

The AO4BPEL \cite{CSHM06} deployment descriptor (DD) is used for specifying non-functional requirements for BPEL processes. First the service (one of logging, security, transactions and reliable messaging) is selected. Then the class (e.g. authentication or con-fidentiality for security) can be specified. The type (e.g. encryption or decryption for confi-dentiality) is then the concrete action that must be executed in order to enforce the re-quirement. Then a selector can be defined such as for a certain BPEL activity. Each selector can be mapped to a requirement.

The W3C proposes a set of languages and protocols for specifying the NFP of a ser-vices’ coordination. A classification of NFP protocols, provided by the WS-* family protocols proposed by the World Wide Web Consortium (W3C). 
%These protocols suppose an underlying communication protocol. 
%
%The following sections describe representative examples of the WS-* protocol-based approaches.
%
 Standards in services' composition implement functional, non-functional constraints and communication aspects by combining different languages and protocols. WSDL and SOAP are languages used respectively for describing services' interfaces and message exchange protocols for calling methods exported by such services. For adding a transactional behaviour to a services' composition it is necessary to implement WS-Coordination, WS-Transaction, WS-BussinessActivity and WS-AtomicTransaction. The selection of the adequate protocols for adding a specific non-functional constraints to a services' composition (e.g., security, transactional behaviour and adaptability) is responsibility of a programmer. As a consequence, the development of an application based on a services' composition is a complex and a time-consuming process.
 % This is opposed to the philosophy of services that aims at facilitating the integration of distributed applications. 
 Other works, like \cite{Fauvet05} introduce a model for transactional services composition based on an advanced transactional model.\cite{BhiriGP05} proposes an approach that consists of a set of algorithms and rules to assist designers to compose transactional services. In \cite{Vidyasankar:2004} the model introduced in \cite{SchuldtABS02} is extended to web services for addressing atomicity.

@@

\subsection{Non-functional properties in service compositions}

Several approaches adding non-functional properties to service compositions extend the underlying composition model. Non-functional properties can concern for example securty, atomicity, availability, response time.
\cite{CholletL09,SchmelingCM11}  associate non-functional  properties  to functional activities  modeled by a web service composition model. In \cite{chmelingCM11} 
non-functional propoerties are  are modeled independently and then associated to   a specific service composition represented in a BPMN diagram. 
In \cite{OvaskaEHPA10} non-functional properties concern  quality management and they are specified at design time. Quality requirements  are then associated to the components of an architecture that implements a service composition.

Other approaches specify architectures or frameworks to manage and validate QoS attributes in service composition processes. 
\cite{XiaoCZBOLH08} proposes a framework to verify NFP compliance in composed services. Tasks in a BPEL process can be annotated with non-functional attributes (such as time, cost, resources, etc.) which are validated at design time. 
\cite{SchmelingCM11} uses a formal method  to choose best suited services instances according to a service composition pattern that includes some QoS attributes. 
 \cite{Babamir2010} proposes a framework for evaluating quality attributes of web services. By executing an algorithm, the framework allow to choose the best web services for any task in a process and also candidate web service to replace original ones in cases it cannot perform at runtime. \cite{DAmbrogio06,Liu20121080}  extend  WSDL and BPEL   for includin  QoS attributes that can be evaluated at runtime.
 \cite{Karunamurthy2012787} proposes an architecture in which web services are described with non-functional characteristics that are validated in the service composition process.

Other approaches \cite{JeongCL09,Kamalabad2012,MohantyRP10} use matching techniques for choosing services that can cope to non-functional properties required by a given application.
\cite{AgarwalLS09} defines a policy language to specificy constraints on functional  properties of Web services for facilitating automated discovery and selection of services. 
\cite{Yeom2006} define a quality driven selection of web services providing a QoS model.

%\cite{} works also on the selection of web service for use considering non-functional properties. The paper present web services quality prediction models, which take non-functional properties into account.

Our proposal in contrast, focuses on the design of service-oriented application with non-functional properties modelled  for services and service compositions and then ensured at execution thanks to appropriate environments.

%We propose to model non-functional properties  represented in our work as constraints and policies but more general quality model such as the proposed by \cite{Goeb2011,Klass2009} could be taken in consideration in further works.

@@

 \cite{Ramollari_asurvey,PapazoglouH06,FeuerlichtM05,soma,Arsanjani:2008} propose methodologies for building service-based applications that  mainly focus on modeling functional aspectos of service-based business processes. SOD-M \cite{decastro1} defines a model-driven approach for development service-based applications. 
Few service-oriented development methodologies    explicitly address  non functional properties modeling \cite{Tran2012531,GutierrezRF10}. \cite{GutierrezRF10} presents PWSSec process, a methodology that provides  activities, tasks, tools, security artifacts and organizational structures necessary to design a secure WS-based solution.

@@

%Besides the notion of model at each level of abstraction, MDA propose the use of \textit{model transformations} within and between levels. Intra-level transformations are used to provide a unified representation of concepts of a given level. Inter-level transformation implement a refinement process between levels. Transformations may be automatic or semi-automatic.

%The methodology for supporting the construction of service-oriented applications, taking into account both functional and non-functional requirements presented in this work extends the \textit{Service Oriented Development Method} (SOD-M) proposed by De Castro et al.\cite{decastro1}.
